The autocorrelation functions in cryo--EM

### 2018-05-24 TEX/The autocorrelation functions in cryoEM.tex
the autocorrelation function of cryo-EM

\title{Toward single particle reconstruction without particle picking: Breaking the detection limit}

@@

\begin{abstract}
	Here comes the abstract
\end{abstract}

\section{Introduction}

\TODO{Revise--Cryo--electron microscopy (cryo--EM) is an innovative technology for single particle reconstruction (SPR) of macromolecules.} 
%In recent years, structures of many molecules, previously regarded as insurmountable, are now being reconstructed to near-atomic resolution
%; see for instance~\cite{kuhlbrandt2014resolution,bartesaghi20152,smith2014beyond}.
%This technological advancement was recognized by the 2017 Nobel Prize in Chemistry~\cite{nobel}. 
% 
In a cryo--EM experiment, biological samples are rapidly frozen in a thin layer of vitreous ice. Within the ice, the molecules are randomly oriented and positioned. The microscope produces a 2D image of the samples embedded in the ice called a \emph{micrograph}. Each micrograph contains tomographic projections of the samples at unknown locations and under unknown viewing directions. The goal is to construct 3-D models of the molecules from the micrographs.

The signal to noise ratio ($\SNR$) of the projections in the micrographs is a function of two dominating factors. On the one hand, the $\SNR$ is a function of the electron dose. To keep radiation damage within acceptable bounds, the dose must be kept low, which leads to high noise levels. On the other hand, the $\SNR$ is a function of the molecule size. The smaller the molecules, the fewer detected electrons carry information about them.

%\paragraph{Particle picking.}
All contemporary methods in the field split the reconstruction procedure in several stages.
The first stage consists in extracting the various particle projections from the micrographs. This is called \emph{particle picking}. Later stages aim to construct a 3-D model of the molecule from these projections. The quality of the reconstruction eventually hinges on the quality of the particle picking stage. As can be seen from Figure~\ref{fig:micro_example}, locating the particles becomes increasingly challenging as the $\SNR$ degrades.

Crucially, it can be shown that reliable detection of individual particles is impossible below a certain critical $\SNR$. This fact has been recognized early on by the cryo-EM community. In particular, in a highly cited paper from 1995, Henderson~\cite{henderson1995limitations} investigates the following questions:
\begin{quote}
	\emph{For the purposes of this review, I would like to ask the question: what is the smallest size of free-standing molecule whose structure can in principle be determined by phase-contrast electron microscopy? Given what has already been demonstrated in published work, this reduces to the question: what is the smallest size of molecule for which it is possible to determine from images of unstained molecules the five parameters needed to define accurately its orientation (three parameters) and position (two parameters) so that averaging can be performed?}
\end{quote}
In that paper and in others that followed (e.g.,~\cite{glaeser1999electron}), it was established that particle picking is impossible for molecules below a certain weight.
%This has often been understood as meaning that particle reconstruction itself is impossible for such molecules~\cite{khoshouei2017cryo,liu2018nearatomic,liang2017phase}. \TODO{This list may be unfair, as some of these papers (in particular, \cite{liang2017phase}) just state that small molecules are very challenging, which is true.}
As a result, it is impossible to reconstruct such small molecules by any of the existing computational pipelines for single particle analysis, as the particles themselves cannot be picked from the micrographs.
%
%\TODO{either here, or in a close neighborhood of this paragraph you should mention the advances that the field has made in recent years towards reconstruction of small size molecules: mention Volta phase plate and the scaffolding cage (and cite the papers I sent you).}
This has motivated recent technical advances in the field, including the use of Volta phase plates~\cite{khoshouei2017cryo,liang2017phase} and scaffolding cages~\cite{liu2018nearatomic}.
%
%
Despite this progress, detecting small molecules in the micrographs remains a challenge.

In this paper, we argue that there is a gap between the two questions in the quoted excerpt above, and that one may be able to exploit it to design better reconstruction algorithms.
Specifically, the impossibility of particle picking does not necessarily imply impossibility of particle reconstruction.
%This can be understood from a statistical point of view by counting the number of parameters that one aims to estimate. \TODO{For a given target resolution, the number of parameters required to describe the 3-D model of the molecule is fixed: it does not depend on the amount of data collected. In contrast, particle picking involves estimation of several so-called nuisance parameters for each projection present in the micrographs, such as their locations for example. To overcome low $\SNR$s, the ratio of data to parameters must increase. This is impossible if the number of parameters grows with the amount of data collected, but not so if the number of parameters remains fixed. Simply put, the key is to bypass the particle picking stage.}
%%Thus, acquiring more data does not necessarily allow to improve the particle picking stage, whereas it should in principle allow to improve the molecule reconstruction.
%%Indeed, the signal itself (the many different views of the molecule) is still present in the micrograph: 
%Indeed, in principle at least, the micrograph does contain the necessary information (several views of the molecule(s) of interest). While we may not be able to locate these views, we may still be able to extract the relevant information through other means.
Indeed, the aim is only to reconstruct the molecule: estimating the locations of the particles in the micrograph is merely a helpful intermediate stage when it can be done. Our main message is that the limits particle packing imposes on molecule size do not translate into limits on particle reconstruction.

%\TODO{Paragraph about the simplified model}
As a proof of concept \TODO{To support our point?}, we study a simplified model where an unknown image appears numerous times at unknown locations in several micrographs, each affected by additive Gaussian noise---see Figure~\ref{fig:micro_example} for an illustration.
The goal is to estimate the planted image. The task is interesting in particular when the $\SNR$ is low enough that particle picking cannot be done reliably.
%Figure~\ref{fig:Einst_example} displays the results of our method, detailed in Section~\ref{sec:results}.
A precise mathematical formulation of the model, including an extension where more than one planted images are to be recovered, is provided in Section~\ref{sec:methods}.
To be clear, we do not consider here many prominent features of real SPR experiments and do not aim to reconstruct any 3-D structure. Instead, we solve a simpler problem that we believe captures key elements of the SPR problem. We note that similar models emerge in spike sorting~\cite{lewicki1998review}, passive radar~\cite{gogineni2017passive} and system identification~\cite{ljung1998system}.

In order to recover the planted image, we use autocorrelation analysis. In a nutshell, we relate the autocorrelation functions of the micrographs to the autocorrelation functions of the planted image.
%To do so, we assume the image occurrences are well separated.
For any noise level, these autocorrelations can be estimated to any desired accuracy provided individual occurrences of the image are well separated and the image appears sufficiently many times in the micrographs. Importantly, there is no need to detect individual occurrences. The autocorrelations of the micrographs are straightforward to compute and require only one pass over the data. These directly yield estimates for the autocorrelations of the image. To estimate the image itself from its estimated autocorrelations, we solve a nonlinear inverse problem via least-squares or a phase retrieval algorithm; see Figure~\ref{fig:Einst_example} for an illustration and Section~\ref{sec:methods} for details. As a side note, we mention that expectation-maximization (EM)---a popular framework in SPR---is intractable for this problem; see Appendix~\ref{sec:theory} for a discussion.

Another interesting feature of the described approach pertains to model bias, whose importance in cryo-EM was stressed by a number of authors~\cite{shatsky2009method,henderson2013avoiding,vanheel2013finding}. In the classical ``Einstein from noise'' experiment, multiple realizations of pure noise are aligned to a picture of Einstein using cross-correlation and then averaged. In~\cite{shatsky2009method}, it was shown that the averaged noise rapidly becomes remarkably similar to the Einstein template. In the context of cryo--EM, this experiment exemplifies how prior assumptions about the particles may influence the reconstructed structure. This model bias is common to all particle picking methods based on template matching. In our approach, no templates are required.

\TODO{What about this one? \cite{vanheel1992correlation}}

%\TODO{Paragraph about bias; can also mention computational aspects (single pass, etc.)}

%\TODO{Already mention challenges? I think it's a good place.} -- no, not here: in conclusion

\clearpage

%\begin{figure}[h!]
%	\centering
%	\begin{subfigure}[h]{0.3\textwidth}
%		\centering
%		\includegraphics[scale=0.3]{Einstein}
%		\caption{\label{fig:Einstein}Einstein image (template)}
%	\end{subfigure}%
%	\begin{subfigure}[h]{0.3\textwidth}
%		\centering
%		\includegraphics[scale=0.3]{Einstein_from_noise_n10}
%		\caption{$n = 10$}
%	\end{subfigure}
%	\begin{subfigure}[h]{0.3\textwidth}
%		\centering
%		\includegraphics[scale=0.3]{Einstein_from_noise_n100}
%		\caption{$n = 10^2$}
%	\end{subfigure}

%	\begin{subfigure}[h]{0.3\textwidth}
%	\centering
%	\includegraphics[scale=0.3]{Einstein_from_noise_n1000}
%	\caption{$n = 10^3$}
%\end{subfigure}%
%\begin{subfigure}[h]{0.3\textwidth}
%	\centering
%	\includegraphics[scale=0.3]{Einstein_from_noise_n10000}
%	\caption{$n = 10^4$}
%\end{subfigure}
%\begin{subfigure}[h]{0.3\textwidth}
%	\centering
%	\includegraphics[scale=0.3]{Einstein_from_noise_n100000}
%	\caption{$n = 10^5$}
%\end{subfigure}

%\caption{\label{fig:EinsteinFromNoise} Einstein from noise experiment. As a template, we used Einstein's image of size $50\times 50$ pixels. We drawn $n$ images of the same size whose entries are composed of i.i.d.\ Gaussian entries with the same mean and variance as Einstein's image. Each noise image was aligned with respect to the template by the maximal value of their cross correlation. After the alignment, all the images were averaged. As can be seen, the more pure noise images we use, the closer the averaged image to the template.   }	
%\end{figure}

%Besides the model bias, particle pickers have several limitations due to the high noise level, in particular for small particles and in low contrast. 
%As a result, the projections in the 2D images are typically not centered; including the correct centers as parameters dramatically increases the complexity of reconstruction algorithms.
%In addition, the information from particles that are too close to each other is usually neglected. Hence, valuable information that can be harnessed is omitted.

%\paragraph{Simplified model for SPR}

%In this paper, we present a preliminary study that aims to show that estimating a particle 
%directly from the micrograph, without identifying the location of individual particle images in the micrograph, might be possible.

%In particular, we consider the toy problem of estimating a set of signals $x_1,\ldots,x_K$ from their multiple occurrences in unknown  locations in a noisy data sequence $y$. Here, $y$ plays the role of the micrograph and the $K$ signals can be thought of as $K$ different viewing directions of the particle, each of which  appears multiple times in the micrograph. 
%A precise mathematical formulation of the model and the estimation problem is provided in Section~\ref{sec:methods}.
%To be clear, we do not consider here many prominent features of real SPR experiments and do not aim to reconstruct any 3-D structure; instead, we solve a simpler problem that we believe captures key elements of the true SPR problem. We also mention in passing that similar models emerge in different scientific fields, such as spike sorting~\cite{lewicki1998review}, passive radar~\cite{gogineni2017passive} and system identification~\cite{ljung1998system}.

\begin{figure}[h!]
	\centering
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_clean}
		\caption{$\sigma = 0$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s05}
		\caption{$\sigma = 0.5$}
	\end{subfigure}
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s3}
		\caption{$\sigma = 3$}
	\end{subfigure}
	\caption{\label{fig:micro_example} Example of micrographs of size $250\times 250$ with additive white Gaussian noise of variance $\sigma^2$ for increasing values of $\sigma$. Each micrograph contains four occurrences of the same $50 \times 50$ image of Einstein. In panel (c), the noise level is such that it is very challenging to locate the occurrences of the planted image. In fact, it can be shown that at low $\SNR$, reliable detection of individual image occurrences is impossible, even if the true image is known. By analogy to cryo-EM, this depicts a scenario where particle picking cannot be done.}	
\end{figure}

%\paragraph{Autocorrelation analysis.}
%We now return to the general problem of estimating $x_1,\dots,x_K$ from the noisy observation $y$.
%In order to recover the signals $x_1,\dots,x_K$ from the data, we use autocorrelation analysis.
%In a nutshell, the method consists of two stages. First, we estimate a mixture (i.e., linear combination) of the low--order autocorrelation functions of the signals from the data. These quantities can be estimated, to any desired accuracy, if individual occurrences are separated by the maximum signal size and each signal appears sufficiently many times in the data. There is no need to detect individual occurrences.
%In the second stage, the signals are estimated from the mixed autocorrelations using a nonconvex least-squares (LS) algorithm or a phase retrieval algorithm; see Section~\ref{sec:methods} for details.
%This method requires only one pass over the data and can recover the signals in any $\SNR$, if the signals appear enough times in the data. As a side note, we mention that expectation-maximization---a popular framework in SPR---is intractable for this problem (see Appendix~\ref{sec:theory}  for more details).

%\paragraph{Model bias.} The significance of the model bias in cryo--EM was stressed in~\cite{shatsky2009method,henderson2013avoiding} and demonstrated by the ``Einstein from noise'' example.
%In this experiment, the image of Einstein is correlated with multiple images of pure noise. The noisy images are aligned to Einstein's image using cross-correlation and then averaged. 
%In~\cite{shatsky2009method}, it was shown that with merely  1000 noise images, the averaged image is remarkably similar to the template image of Einstein rather than to pure noise.
%In Figure~\ref{fig:EinsteinFromNoise} we conducted this experiment with different number of noise images. With merely 1000 noise images, Einstein's face is clearly observed in the averaged image. With $10,000$  image, the output is remarkably similar to the template image of Einstein rather than to pure noise. Crucially, the more noisy images we use, the more severe the model bias towards the template. 
%In the context of cryo--EM, this example demonstrates how our prior assumptions about the particle may influence the reconstructed structure.
%This true in general to any method which is based on template matching.
%\TODO{Do we have a nice experiment to exemplify the phenomenon?}

\section{Results} \label{sec:results}

In this section we describe  some numerical experiments and show results. The experimental details and the algorithms are discussed at length in Section~\ref{sec:methods}.

In the first experiment, we estimated Einstein's image of size $50\times 50$ from multiple micrographs of size $4096\times 4096$ pixels, contaminated  with additive i.i.d.\ Gaussian noise with standard deviation $\sigma=3$ ($\SNR=1/20$). An illustrative example of a micrograph appears in the right panel of Figure~\ref{fig:micro_example}.
Each micrograph contains multiple occurrences of Einstein's image  in random locations, while keeping a separation of  $49$ pixels between the images in each direction. From the micrographs, we estimated the signal's power spectrum, which is equivalent to the second-order autocorrelation. Then, in order to estimate the signal itself, we applied a standard phase retrieval algorithm called relaxed-reflect-reflect (RRR).

As an initial guess, we picked an image of the physicist Issac Newton; see Figure~\ref{fig:newton}. If the algorithm was prone to model bias, we would expect to get as an output an image that resembles Newton, similarly to the ``Einstein from noise'' effect of Figure~\ref{fig:EinsteinFromNoise}. However, the experiment exhibits the desired phenomenon: the more data we collect, the better the reconstruction quality.

Figure~\ref{fig:Einst_example} demonstrates several recovery results for different number of recorded micrographs. Figure~\ref{fig:error_per_micro} presents the normalized recovery error and the power spectrum estimation error as a function of the number of micrographs.
To measure the error, we use the root mean square error (RMSE) defined as 
\begin{equation}
\text{RMSE}  := \frac{\|x - \hat{x}\|_{\text{F}}}{\|x \|_{\text{F}}},
\end{equation} 
where $x$ and $\hat{x}$ are, respectively, the underlying and estimated image, and $\|\cdot\|_{\text{F}}$ stands for the Frobenius norm. 
As expected, the RMSE of estimating the power spectrum   decreases linearly with slope $-1/2$ in logarithmic scale. 
\TODO{Few comments: 1. We have a movie in the supplementary material. 2. Does the error also have the right slope? 3. We can put more images to exemplify the progress 4. Need to improve the micrograph's generation code  }

\begin{figure}[h!]
	\centering
	\begin{subfigure}[h]{0.25\textwidth}
		\centering
		\includegraphics[scale=0.3]{Newton}
		\caption{\label{fig:newton}Newton  (model)}
	\end{subfigure}%
	\begin{subfigure}[h]{0.25\textwidth}
		\centering
		\includegraphics[scale=0.3]{reconstruction2}
		\caption{\small $P =1024$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.25\textwidth}
		\centering
		\includegraphics[scale=0.3]{reconstruction20}
		\caption{$P = 1024\times 10$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.25\textwidth}
		\centering
		\includegraphics[scale=0.3]{reconstruction200}
		\caption{$P = 1024\times 100$}
	\end{subfigure}%
	\caption{\label{fig:Einst_example} \TODO{Caption and image not yet revised} Recovery of Einstein's image. The RRR algorithm is initialized with the image of the physicist Issac Newton in the left panel. We show estimations of Einstein using $P=1024,1024\times 10$ and $1024\times 100$ number of micrographs, each contains $700$ image occurrences on average and the $\SNR$ is $1/20$. The signal's estimation RMSE is $0.5514$, $0.3494$ and $0.2112$, respectively. The power spectrum estimation RMSE is, respectively, $0.0269 $, $0.0083$ and $0.0027$. \TODO{Can add more images to exemplify the progress}\TODO{Resolution of Newton too high?}}	
\end{figure}

%
%Figure~\ref{fig:1Dheterosignals} demonstrates accurate estimation of three signals simultaneously with noise level $\sigma=3$. We define the ratio of the space occupied by the $i$th signal as 
%\begin{equation}
%\gamma_i = \frac{M_i L}{N},
%\end{equation}
%where $M_i$ is the signal's number of occurrences, $L$ is the length of the signal and $N$ is the micrograph length. In the experiment, we do not assume to know these ratios, neither the noise level $\sigma$. As can be seen, given enough signals occurrences, we can estimate accurately the signals. he estimation quality 
%is poorer than the other two signals, a phenomenon that can be explained using Proposition~\ref{prop:uniqueness}.

Results from a similar experiment with three ($K=3$) one-dimensional signals are presented in Figure~\ref{fig:1Dheterosignals}.
 We define the ratio of the space occupied by the $i$th signal as 
\begin{equation}
\gamma_i = \frac{M_i L}{N},
\end{equation}
where $M_i$ is the number of signal's  occurrences, $L$ is the length of the signal and $N$ is the micrograph length.
These ratios present the {density} of the the signals in the data.
In the experiment, we do not assume to know these ratios, neither the noise level $\sigma$.

In order to estimate the signals, we computed the first three autocorrelation functions of the data and then estimated the signals and their corresponding $\gamma_i$ using a nonconvex least-squares. 
As can be seen, given enough signal occurrences, we can estimate accurately the signals. 
The estimation quality of the triangle signal is poorer than the other two signals, a phenomenon that is explained using Proposition~\ref{prop:uniqueness}.
\TODO{1. We will replace the figure with a ``progress figure'' (like we did for Einstien) for each signals and a plot of the recovery error for all three signals. 2. We may want to replace the triangle signal}
%We do not assume the knowledge of the noise level. 
%Crucially, starting from the point of $n\approx10^8$, the RMSE of the signal estimation decreases linearly with slope $-1/2$---the expected estimation rate of the autocorrelation functions---implying the stability of the recovery algorithm.

\begin{figure}[h]
\centering
\includegraphics[scale=.7]{Einstein_progress}
\caption{\label{fig:error_per_micro} The left panel shows the RMSE of estimating the power spectrum of the signal from the data as a function of the number of  collected micrographs. As expected, the curve is linear with slope $-1/2$ in logarithmic scale. The right panel shows the RMSE  (in logarithmic scale) of recovering Einstein's image. \TODO{The program still runs so these figures will be extended to the right (more micrographs).}}	
\end{figure}

%
%\begin{figure}[t]
%	\centering
%	\includegraphics[scale=0.45]{progressive_n}
%	\caption{\TODO{...} \TODO{This figure is with new ROI method based on loss functions}}
%	\label{fig:1Dhomosignals}
%\end{figure}
%
%\begin{figure}[t]
%	\centering
%	\includegraphics[width=0.5\linewidth]{XP_1D_homogeneous_new_ROI/progressive_RMSE_n}
%	\caption{\TODO{...} \TODO{This figure is with new ROI method based on loss functions}}
%	\label{fig:1DhomoRMSE}
%\end{figure}

\begin{figure}[t]
	\centering
	\includegraphics[width=\linewidth]{XP_1D_heterogeneous/hetero_example}
	\caption{An experiment for estimate three one-dimensional signals simultaneously with noise level $\sigma=3$. Red signals are the ground truth (targets) and the blue signals are our estimations. Individual RMSE of the estimates: $0.0239393 / 0.208925 / 0.0335956$. The estimated $\gamma$'s are $0.02574 / 0.01693 / 0.00818$ and the true ones $0.02561 / 0.01707 / 0.00853$. The $\SNR$ is $1/12$. \TODO{1. to replace with a ``progress'' plot 2. replace the triangle signal 3. add error progress figure}}
	\label{fig:1Dheterosignals}
\end{figure}

\section{Discussion}

All current algorithmic pipeline for SPR using cryo--EM start with a particle picking algorithm which is prone to model bias. 
Bypassing the particle picking stage and constructing a 3-D model directly from the data---without assuming prior knowledge on the particle to be estimated---can be used to reconstruct ab initio models to initialize a refinement algorithm.  Alternatively, it can be applied to generate templates for a particle picker which does not suffer from model bias.

In this paper, we examined a simplified model that, we believe, captures important features of the SPR problem. 
While the field is currently dominated by Bayesian methods such as EM, they are intractable for such problems. As an alternative, we propose to use autocorrelation analysis technique that shares some common lines (did you get the wordplay?)  with Kam's method for ab inito modeling~\cite{kam1980reconstruction,levin20173d,singer2018mathematics}. 
That being said, the SPR model is far more complicated than the model presented here. In a future research, we hope to bridge this gap.

Our results rely on two core assumptions that are not necessarily met by an SPR experiment.  
First, we modeled the background information as i.i.d.\ additive noise. In practice, the background information may be structured or signal dependent. 
Second, we assumed that the signal occurrences are all separated by the length of the signal, see~\eqref{eq:spacing}. 
This separation can be induced by careful experimental design \TODO{???}. 
If the signals are not separated, one can introduce a new variable that represents the distribution of the spacing between signal occurrences and then compute explicitly the relation between the autocorrelation functions of the data and the signals. It is yet to be studied under what conditions on this distribution one can estimate the signals.

%The first entry $p[1]$ will represent the probability that two consecutive signals are separated by only one entry, $p[2]$ the probability for spacing of two entries and so on. Using this auxiliary variable $p$, one can write explicitly the relation between the autocorrelation functions of the data and those of the signal in a similar way to~\eqref{eq:data_ac}. 
%An interesting question is  under what conditions on $p$ and the signals, one can estimate the signals from the data.

\section{Methods} \label{sec:methods}

\paragraph{Mathematical model.}

Let $x_1,\ldots,x_K\in\RL$ be the sought signals and let $y\in\RN$ be the data. 
The forward model can be posed as a mixture of \emph{blind deconvolution} problems between binary signals and the target signals $x_i$:
\begin{equation} \label{eq:model}
y = \sum_{i=1}^K x_i\ast s_i + \varepsilon,\quad \varepsilon\sim\mathcal{N}(0,\sigma^2 I).
\end{equation}
%We model the background information as i.i.d.\ Gaussian noise with zero mean and $\sigma^2$ variance. 
The nonzero values of each  $s_i\in\{0,1\}^N$ determine the position of $x_i$'s occurrences. 
We denote the set of these nonzero values by 
 $\mathcal{S}_i$ and its cardinality by $\vert \mathcal{S}_i\vert = M_i$. 
By assuming that all $\mathcal{S}_i$'s are disjoint,  we let $s = \sum_{i=1}^Ks_i$, $\mathcal{S} = \bigcup_{i=1}^{K} \mathcal{S}_i$ and  $\vert \mathcal{S}\vert :=M =  \sum_{i=1}^{K}M_i$.  %Neither the $M_i$'s nor $M$ are assumed to be known.
 Literature survey on blind deconvolution and related problems is given in Appendix~\ref{sec:related_literature}.

In order to estimate the mixture of autocorrelations, we assume that the support of $s$ is not clustered. In particular, we assume that
\begin{equation} \label{eq:spacing}
  \vert i-j \vert\geq L-1,   \quad \text{for all } i,j\in\mathcal{S} \text{ such that } i\neq j.
\end{equation}
The goal of the problem is to estimate $x_1,\ldots,x_K$ from $y$.

%
%\section{Autocorrelation analysis}   \label{sec:autocorrelation}
%
%Our method for estimating the signals is composed of two stages. 
%First, we use the autocorrelation functions of the data to estimate a mixture (i.e., linear combination) of the $K$ signal's autocorrelations. The mixed autocorrelation can be estimated to any accuracy, in any $\SNR$ level, if $M$ is large enough and the spacing condition~\eqref{eq:spacing} is met. Then, we  use a nonconvex LS  to estimate the signals from their mixed autocorrelations. 
%In this section, we elaborate on the autocorrelation functions and their estimations, while the precise recovery procedure, based on nonconvex optimization, will be discussed in detail in the next section.

\paragraph{Aperiodic autocorrelation functions}
%\subsection{Aperiodic autocorrelation functions} \label{sec:aperiodic_ac}
For  
$z\in\RL$ and $k\geq 2$, the autocorrelation of order $k$ is defined for any integer shifts $\ell_1, \ldots, \ell_{k-1}$ by
\begin{align}
	a_z^k[\ell_1,\ldots,\ell_{k-1}]  & = \sum_{i=-\infty}^{+\infty} z[i]z[i+\ell_1]\ldots z[i+\ell_{k-1}],
	\label{eq:ac_general}
\end{align}
where indexing of $z$ out of the bounds $0, \ldots, L-1$ is zero-padded, as usual.
Explicitly, the mean (which can be thought of as a first-order autocorrelation) and the second and third autocorrelations are given by
\begin{align} 
	a_z^1 & = \sum_{i=0}^{L-1} z[i], \nonumber\\
	a_z^2[\ell] & = \sum_{i = \max\{0, -\ell\}}^{L-1 + \min\{0, -\ell\}} z[i]z[i+\ell], \nonumber\\
	a_z^3[\ell_1,\ell_2] & = \sum_{i = \max\{0, -\ell_1, -\ell_2\}}^{L-1 + \min\{0, -\ell_1, -\ell_2\}} z[i]z[i+\ell_1]z[i+\ell_2]. \label{eq:ac_special}
\end{align}
Note that the autocorrelation functions are symmetric so that $a_z^2[\ell] = a_z^2[-\ell]$ and also $$a_z^3[\ell_1,\ell_2] = a_z^3[\ell_2,\ell_1]=a_z^3[-\ell_1,\ell_2-\ell_1].$$
%Additionally, if the moments of the signal depend only on the difference between the indices (Toeplitz structure), then they are equivalent to the autocorrelation functions.

\paragraph{Estimating the autocorrelation function of a single signal.}

We first consider the problem of estimating the autocorrelations of a single signal from the data. 
The main principles carry through for $K>1$ as will be shown next.

%In order to estimate the autocorrelations of the signal, we first compute the first $L$ entries of the data's autocorrelations. 
For the purpose of the analysis, we consider  the asymptotic regime where $M,N\to\infty$, while preserving fixed ratio. 
Specifically, we define the ratio of the measurement occupied by  the signal as
\begin{equation}
\gamma = \frac{M L}{N}.
\end{equation}
Under the spacing constraint~\eqref{eq:spacing}, we have $\gamma\leq\frac{L}{2L-1}\approx 1/2$.

The main pillar of this work is the following simple observation.
If the support signal $s$ satisfies the spacing constraint~\eqref{eq:spacing}, then the first $L$ entries of the data autocorrelations converge 
to a scaled, biased, version of the signal's autocorrelation:
\begin{align} 
\lim_{N\to\infty} a_y^1 & = \gamma a_{x}^1, \nonumber\\
\lim_{N\to\infty} a_y^2[\ell] & = \gamma a_{x}^2[\ell] +\sigma^2\delta[\ell], \label{eq:data_ac_k1} \\
\lim_{N\to\infty} a_y^3[\ell_1,\ell_2] & = \gamma a_{x}^3[\ell_1,\ell_2] + \sigma^2\gamma a_{x}^1 \big(\delta[\ell_1,0]+\delta[0,\ell_2]+\delta[\ell_1,\ell_2]\big), \nonumber
\end{align}
for $\ell,\ell_1,\ell_2=0,\ldots L-1$, and where $\delta$ denotes the Kronecker delta function. 
%These relations are proven in Appendix~\ref{sec:autocorrelation_computation}.

\paragraph{Estimating the autocorrelation function of a multiple signals.}

As before, we consider  the asymptomatic regime where $M_1,\ldots,M_K,N\to\infty$, while preserving fixed ratios
\begin{align}
	\gamma_k = \frac{M_k L}{N}, \quad \gamma = \sum_{k=1}^K\gamma_k.
\end{align}
If the support $s$ satisfies the spacing constraint~\eqref{eq:spacing}, then  similarly to~\eqref{eq:data_ac_k1} one can estimate the mixture of the $K$ signals'  autocorrelations:
\begin{align}
\lim_{N\to\infty} a_y^1 & = \sum_{k=1}^K\gamma_k a_{x_k}^1, \nonumber\\
\lim_{N\to\infty} a_y^2[\ell] & = \sum_{k=1}^K\gamma_k a_{x_k}^2[\ell] +\sigma^2\delta[\ell],  \label{eq:data_ac}\\
\lim_{N\to\infty} a_y^3[\ell_1,\ell_2] & = \sum_{k=1}^K\gamma_k a_{x_k}^3[\ell_1,\ell_2] + \sigma^2\left(\sum_{k=1}^K\gamma_k a_{x_k}^1\right)(\delta[\ell_1,0]+\delta[0,\ell_2]+\delta[\ell_1,\ell_2]), \nonumber
\end{align}
This relation is proven in  Appendix~\ref{sec:autocorrelation_computation}.

\paragraph{Numerical experiments details.}
For the 2-D experiment shown in Figures~\ref{fig:Einst_example} and~\ref{fig:error_per_micro}, we generated $P$ micrographs of size $4096\times 4096$ pixels. 
In each micrograph, we placed Einstein's image of size $50\times 50$  in random locations, while keeping the separation condition~\eqref{eq:spacing}.  
This is done by randomly selecting $4000$ placements in the micrograph, one at a time with
an accept/reject rule based on the separation constraint and locations picked so far. In average, $700$ images were placed in each micrograph.   
Then, an i.i.d.\ Gaussian noise with standard deviation $\sigma=3$ was added inducing $\SNR$ of approximately $1/20$. An example of a micrograph's excerpt is presented in the right panel of Figure~\ref{fig:micro_example}.
%Different micrographs are handled sequentially on a GPU, as GPUs are particularly well suited to execute simple instructions across large vectors of data. If multiple GPUs are available, segments can of course be handled in parallel.

>>>>

	\title{The autocorrelation functions in cryo--EM}

@@

The 3-D Fourier transform of an L-bandlimited 3-D volume (e.g., particle) can be expanded by  spherical harmonics:
\begin{equation} \label{eq:volume}
\hat{V}(k,\theta,\phi)  = \sum_{\ell=0}^{L} \sum_{m=-\ell}^{\ell} A_{\ell,m}(k)Y_\ell^{m}(\theta,\phi),
\end{equation}
where $(\theta,\phi)$ are two angles on the sphere, $k$ is the radial coordinate, $Y_{\ell}^m(\theta,\phi)$ is the spherical harmonic of degree $\ell$ and order $m$ and  $A_{\ell,m}(k)$ are the associated spherical harmonics coefficients, to be estimated. 
A rotation of the volume by $\omega\in SO(3)$ can be described using the Wigner D-function $D_{m,m'}^{\ell}$:
\begin{equation}
\begin{split}
(R_\omega \hat{V})(k,\theta,\phi)  &= \sum_{\ell=0}^{L} \sum_{m=-\ell}^{\ell} A_{\ell,m}(k)(R_\omega Y_\ell^{m})(\theta,\phi) \\
&= \sum_{\ell=0}^{L} \sum_{m=-\ell}^{\ell} A_{\ell,m}(k) 
\sum_{m'=-\ell}^{\ell}{D_{m,m'}^{\ell}(\omega)}Y_\ell^{m'}(\theta,\phi).
\end{split}
\end{equation}

By the Fourier slice theorem, each cryo--EM measurement (projection) is equivalent (through 2-D Fourier transform) to the slice of $\hat{V}$, associated with $\theta= \pi/2$, after $\hat{V}$ was rotated by $\omega\in SO(3)$. Explicitly, the Fourier transform of a projection from the viewing direction $\omega$ is related to the spherical harmonic coefficients of the object through:
\begin{equation} \label{eq:P}
\begin{split}
P_\omega(k,\phi)  = \sum_{\ell=0}^{L} \sum_{m=-\ell}^{\ell} A_{\ell,m}(k) 
\sum_{m'=-\ell}^{\ell}{D_{m,m'}^{\ell}(\omega)}Y_\ell^{m'}(\pi/2,\phi).
\end{split}
\end{equation}

Next, we want to relate the projections $P_\omega$ with the mean and the autocorrelation functions computed directly from the micrograph.
The mean of the micrograph is proptional to 
\begin{equation}
M_1 \propto \sum_{n=1}^N \sum_{x,y}P_n(x,y),
\end{equation}
where $P_n$ denotes the $n$th projection. By taking $n\to\infty$, we get  
\begin{equation}
M_1 \propto \sum_{x,y}\int_{\omega}P_\omega(x,y)\rho(\omega) d\omega,
\end{equation}
where $\rho(\omega)$ denotes the (possibly unknown) viewing direction distribution over $SO(3)$.

We assume the projections are sufficiently separated so that, in the limit $n\to\infty$, the $(\Delta_x,\Delta_y)$ entry of the second-order autocorrelation of the micrograph is proportional to:
\begin{equation}
M_2(\Delta_x,\Delta_y)\propto\sum_{x,y}\int_{\omega} P_\omega(x,y)P_\omega(x+\Delta_x,y+\Delta_y)\rho(\omega)d\omega  + \textrm{bias}.
\end{equation}
 The assumption here is that $(\Delta_x, \Delta_y)$ are small enough so that, in computing the auto-correlation, points $(x, y)$ and $(x+\Delta_x, y+\Delta_y)$ do not touch distinct particles. 
In the same way and under the same conditions, the third moment is given by 
\begin{equation} 
M_3(\Delta_x^1,\Delta_y^1;\Delta_x^2,\Delta_y^2) \propto \sum_{x,y}\int_{\omega} P_\omega(x,y)P_\omega(x+\Delta_x^1,y+\Delta_y^1)P_\omega(x+\Delta_x^2,y+\Delta_y^2)\rho(\omega)d\omega + \textrm{bias}.
\end{equation}

In order to determine the particle, by~\eqref{eq:volume} one needs to estimate order of $L^3$ spherical harmonics coefficients. If the pixel size is proportional to $1/L$ (to match the volume's resolution), then $M_3$ provides order of $L^4$ equations involving triple products of $P_\omega$. 
However, 
since  the in-plane rotation of each particle image is usually uniformly distributed, $M_3$  depends on only three parameters: the length of the vector $(\Delta_x^1,\Delta_y^1)$, the length of the vector  $(\Delta_x^2,\Delta_y^2)$ and the  angle between the two vectors.  Therefore,  $M_3$  provides only $\sim L^3$ equations. 
 Since $P_\omega$ depends (after coordinate transformation) linearly in the spherical harmonic coefficients through~\eqref{eq:P}, this means we have a system of $\sim L^3$ cubic equations in the $\sim L^3$ sought parameters. Importantly, the coefficients of these equations can be estimated from the micrographs directly, without particle picking stage.
